---
title: "Download"
---

## Overview

The download module covers four related tasks:

| Task | Function |
|---|---|
| Download one file safely | `download_url()` |
| Download multiple files | `download_batch()` |
| Download gene annotation | `download_gene_ref()` |
| Download GEO series data | `download_geo()` |

```{r}
library(evanverse)
```

The module is intentionally small. It does not try to replace a full data
management workflow; it gives evanverse users predictable helpers for common
download jobs that appear repeatedly in analysis projects.

## Single-File Downloads

`download_url()` downloads one URL to one destination path.

```{r}
dest <- file.path(tempdir(), "README.md")

download_url(
  url  = "https://raw.githubusercontent.com/tidyverse/ggplot2/main/README.md",
  dest = dest
)
```

By default, an existing destination file is treated as complete and the
download is skipped.

```{r}
download_url(
  url       = "https://raw.githubusercontent.com/tidyverse/ggplot2/main/README.md",
  dest      = dest,
  overwrite = FALSE
)
```

Set `overwrite = TRUE` to download again. With the default `resume = TRUE`,
an existing non-empty destination is used as a resume target.

```{r}
download_url(
  url       = "https://raw.githubusercontent.com/tidyverse/ggplot2/main/README.md",
  dest      = dest,
  overwrite = TRUE,
  resume    = TRUE
)
```

When `resume = FALSE`, the function downloads to a temporary file first and
moves it into place only after success. This avoids leaving a new partial file
at `dest` after a failed download.

```{r}
download_url(
  url       = "https://raw.githubusercontent.com/tidyverse/ggplot2/main/README.md",
  dest      = dest,
  overwrite = TRUE,
  resume    = FALSE
)
```

Use shorter timeouts and fewer retries for interactive work; use longer values
for unattended scripts.

```{r}
download_url(
  url     = "https://example.com/data.csv",
  dest    = file.path(tempdir(), "data.csv"),
  timeout = 60,
  retries = 1
)
```

## Batch Downloads

`download_batch()` downloads multiple URLs into one directory. File names are
derived from the URL path.

```{r}
urls <- c(
  "https://httpbin.org/robots.txt",
  "https://httpbin.org/encoding/utf8"
)

paths <- download_batch(
  urls     = urls,
  dest_dir = file.path(tempdir(), "downloads")
)
```

Existing files are skipped when `overwrite = FALSE`.

```{r}
download_batch(
  urls      = urls,
  dest_dir  = file.path(tempdir(), "downloads"),
  overwrite = FALSE
)
```

The function rejects duplicate destination file names. This matters when two
different URLs end with the same file name.

```{r}
download_batch(
  urls = c(
    "https://host-a.example/data.csv",
    "https://host-b.example/data.csv"
  ),
  dest_dir = tempdir()
)
```

Blank URLs are rejected early so errors point to the input rather than to curl
or file-name derivation.

```{r}
download_batch(
  urls     = c("https://example.com/data.csv", ""),
  dest_dir = tempdir()
)
```

## Gene Annotation

`download_gene_ref()` retrieves a standard Ensembl gene annotation table for
human or mouse using `biomaRt`.

```{r}
human_ref <- download_gene_ref("human")

head(human_ref)
```

The result contains stable columns used elsewhere in evanverse workflows:

| Column | Meaning |
|---|---|
| `ensembl_id` | Ensembl gene ID |
| `symbol` | Gene symbol |
| `entrez_id` | Entrez gene ID |
| `gene_type` | Ensembl biotype |
| `chromosome`, `start`, `end`, `strand` | Genomic location |
| `description` | Ensembl description |
| `species` | Requested species |
| `ensembl_version` | Current Ensembl version when available |
| `download_date` | Local download date |

Use `dest` to save the table as an RDS file. The `.rds` extension is appended
when it is omitted.

```{r}
download_gene_ref(
  species = "mouse",
  dest    = file.path(tempdir(), "mouse_gene_ref")
)
```

`biomaRt` is required for this function. If it is not installed, the function
fails with an installation hint instead of a namespace error.

## GEO Series

`download_geo()` downloads a GEO series with three pieces:

| Output | Contents |
|---|---|
| `gse_object` | GEO expression object from `GEOquery::getGEO()` |
| `supplemental_files` | Downloaded supplemental files, when available |
| `platform_info` | Platform ID and downloaded platform annotation files |

```{r}
geo <- download_geo(
  gse_id   = "GSE121212",
  dest_dir = file.path(tempdir(), "GSE121212")
)

names(geo)
```

GEO accessions must use the `GSE` prefix followed by digits.

```{r}
download_geo("GSE12345", dest_dir = tempdir())
```

`GEOquery` is required for this function. Supplemental and platform downloads
can fail independently of the main series matrix, so downstream code should
check the returned file vectors before reading them.

## Design Notes

- `download_url()` uses temporary files for fresh downloads so failed calls do
  not create a destination file that later calls would treat as complete.
- Resume mode writes directly to the destination because that is the intended
  curl resume behavior.
- `download_batch()` retries only failed files after each attempt.
- `download_batch()` currently does not remove failed destination files because
  `curl::multi_download()` may overwrite a user file that existed before the
  call. Cleanup needs to distinguish files created by the current call from
  files owned by the user.
- External data sources are network-dependent. Tests that hit httpbin, Ensembl,
  or GEO are skipped on CRAN and CI, but should still be run locally when
  validating this module.

