---
title: "Stat"
---

## Overview

The stat module covers five common statistical workflows:

| Task | Function |
|---|---|
| Power and sample-size planning | `stat_power()`, `stat_n()` |
| Two-group comparison | `quick_ttest()` |
| Multi-group comparison | `quick_anova()` |
| Categorical association | `quick_chisq()` |
| Correlation analysis | `quick_cor()` |

```{r}
library(evanverse)
```

The module is designed for quick, explicit statistical checks during analysis.
It does not replace a full statistical modelling workflow; it gives small,
repeatable wrappers around common tests with diagnostics, effect-size summaries,
and plotting methods.

## Power Planning

`stat_power()` computes achieved power from sample size and expected effect
size.

```{r}
power_res <- stat_power(
  n           = 30,
  effect_size = 0.5,
  test        = "t_two"
)

print(power_res)
summary(power_res)
plot(power_res)
```

`stat_n()` computes the required sample size for a target power.

```{r}
n_res <- stat_n(
  power       = 0.8,
  effect_size = 0.5,
  test        = "t_two"
)

print(n_res)
plot(n_res)
```

Supported test families share the same effect-size conventions:

| `test` | Meaning | Effect size |
|---|---|---|
| `"t_two"` | Two-sample t-test | Cohen's d |
| `"t_one"` | One-sample t-test | Cohen's d |
| `"t_paired"` | Paired t-test | Cohen's d |
| `"anova"` | One-way ANOVA | Cohen's f |
| `"proportion"` | One-sample proportion test | Cohen's h |
| `"correlation"` | Correlation test | Pearson r |
| `"chisq"` | Chi-square test | Cohen's w |

ANOVA requires `k`, the number of groups. Chi-square requires `df`, the degrees
of freedom.

```{r}
stat_power(n = 25, effect_size = 0.25, test = "anova", k = 3)
stat_n(power = 0.8, effect_size = 0.3, test = "chisq", df = 2)
```

## Two-Group Comparison

`quick_ttest()` compares exactly two groups for a numeric outcome. With
`method = "auto"`, it checks normality and chooses between Welch's t-test and a
Wilcoxon test.

```{r}
set.seed(42)
df <- data.frame(
  group = rep(c("A", "B"), each = 30),
  value = c(rnorm(30, 5), rnorm(30, 6))
)

tt <- quick_ttest(
  data      = df,
  group_col = "group",
  value_col = "value"
)

print(tt)
summary(tt)
plot(tt)
```

Independent comparisons use Welch's t-test when the parametric path is selected.
Paired comparisons require an ID column.

```{r}
set.seed(7)
paired_df <- data.frame(
  id    = rep(seq_len(20), 2),
  group = rep(c("pre", "post"), each = 20),
  value = c(rnorm(20, 10, 2), rnorm(20, 11, 2))
)

quick_ttest(
  paired_df,
  group_col = "group",
  value_col = "value",
  paired    = TRUE,
  id_col    = "id"
)
```

Constant groups are allowed in the normality diagnostic path. Shapiro-Wilk is
skipped for those groups, and the function continues with a clear message
instead of failing inside `stats::shapiro.test()`.

## Multi-Group Comparison

`quick_anova()` compares two or more groups for a numeric outcome. With
`method = "auto"`, it uses group normality diagnostics and variance checks to
choose among classical ANOVA, Welch ANOVA, and Kruskal-Wallis.

```{r}
set.seed(123)
anova_df <- data.frame(
  group = rep(LETTERS[1:3], each = 40),
  value = rnorm(120, mean = rep(c(0, 0.5, 1.2), each = 40), sd = 1)
)

av <- quick_anova(
  anova_df,
  group_col = "group",
  value_col = "value"
)

print(av)
summary(av)
plot(av)
```

Post-hoc tests can be selected explicitly, or left on `auto`:

| Omnibus method | Default post-hoc |
|---|---|
| `anova` | Tukey HSD |
| `welch` | Pairwise Welch t-tests, BH-adjusted |
| `kruskal` | Pairwise Wilcoxon tests, BH-adjusted |

```{r}
quick_anova(
  anova_df,
  group_col = "group",
  value_col = "value",
  method    = "anova",
  post_hoc  = "tukey"
)
```

Kruskal-Wallis results include epsilon-squared as an effect-size summary.

## Categorical Association

`quick_chisq()` tests association between two categorical variables. The
automatic method uses expected frequencies and table shape.

```{r}
set.seed(123)
chisq_df <- data.frame(
  treatment = sample(c("A", "B", "C"), 100, replace = TRUE),
  response  = sample(c("Success", "Failure"), 100, replace = TRUE)
)

cs <- quick_chisq(
  chisq_df,
  x_col = "treatment",
  y_col = "response"
)

print(cs)
summary(cs)
plot(cs)
```

Method selection:

| Case | Method |
|---|---|
| 2x2 with expected frequency below 5 | Fisher's exact test |
| 2x2 with moderate expected frequencies | Chi-square with Yates correction |
| Larger sparse tables | Chi-square with warning |
| Matched square table, explicit request | McNemar's test |

`method = "mcnemar"` is only for paired or matched categorical outcomes.

## Correlation Analysis

`quick_cor()` computes a correlation matrix, p-values, optional p-value
adjustment, and significant-pair summaries.

```{r}
cor_res <- quick_cor(
  mtcars,
  vars = c("mpg", "hp", "wt", "qsec")
)

print(cor_res)
summary(cor_res)
plot(cor_res, type = "upper", show_sig = TRUE)
```

Use `p_adjust_method` when many pairs are tested.

```{r}
quick_cor(
  mtcars,
  vars            = c("mpg", "hp", "wt", "qsec"),
  method          = "spearman",
  p_adjust_method = "BH"
)
```

The `use` argument follows `stats::cor()` missing-value handling and is
validated before the correlation is computed:

```{r}
quick_cor(
  mtcars,
  use = "pairwise.complete.obs"
)
```

Valid choices are `everything`, `all.obs`, `complete.obs`, `na.or.complete`,
and `pairwise.complete.obs`.

## Design Notes

- `stat_power()` and `stat_n()` return the same S3 class, `power_result`, so
  `print()`, `summary()`, and `plot()` work consistently.
- `quick_ttest()` uses Welch's t-test for independent parametric comparisons.
- Wilcoxon tests use `exact = FALSE` so common tied data do not create noisy
  exact-p-value warnings.
- Normality diagnostics skip constant groups instead of failing inside
  Shapiro-Wilk.
- `quick_anova()` still runs diagnostics when a method is forced, but the
  forced method is respected.
- `quick_chisq()` reports observed counts, expected frequencies, residuals when
  available, and Cramer's V when a chi-square statistic exists.
- `quick_cor()` reports significant pairs based on adjusted p-values when
  `p_adjust_method` is not `"none"`.
- Plot methods return the original result invisibly after printing the ggplot,
  matching the package's reporting style.

