# Data Cleaning Essentials

Data cleaning is one of the most critical steps in any data workflow.\
This chapter outlines reusable strategies for cleaning and preparing your data for analysis.

------------------------------------------------------------------------

## 3.1 Inspecting Data

Use tools like `summary()`, `str()`, or `skimr::skim()` to understand the structure and quality of your dataset.

```{r}
summary(mtcars)
```

## 3.2 Handling Missing Values

Common approaches include:

-   Removing rows with `NA`

-   Imputing values

-   Using domain knowledge to infer missing information

```{r}
#| eval: false
# Drop rows with missing values
clean_data <- na.omit(raw_data)
```

## 3.3 Removing Duplicates

```{r}
#| eval: false
clean_data <- distinct(raw_data)
```

Use `duplicated()`, `distinct()` (dplyr), or `janitor::get_dupes()`.

## 3.4 Standardizing Column Names

```{r}
#| eval: false
library(janitor)
clean_data <- raw_data |> janitor::clean_names()
```

## 3.5 String Cleaning

Use `stringr::str_trim()`, `str_to_lower()`, etc., to normalize text.

```{r}
#| eval: false
library(stringr)
df$name <- str_to_title(str_trim(df$name))
```