# Data Visualization Basics

This chapter introduces the basics of data visualization using **ggplot2**.

> *Visualizations turn raw data into insights.*

------------------------------------------------------------------------

## 4.1 The Grammar of Graphics

The core idea of `ggplot2` is to build plots layer by layer:

```{r}
library(ggplot2)

ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point()
```

## 4.2 Common Plot Types

### 📊 Bar Chart

```{r}
ggplot(mpg, aes(class)) +
  geom_bar()
```

### 📦 Boxplot

```{r}
ggplot(mpg, aes(class, hwy)) +
  geom_boxplot()
```

### 🟢 Scatter Plot

```{r}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point()
```

## 4.3 Themes and Aesthetics

Use `theme_minimal()`, `theme_classic()` and control colors, sizes, and fonts.

```{r}
ggplot(mpg, aes(class, hwy)) +
  geom_boxplot() +
  theme_minimal()
```

## 4.4 Exporting Figures

Save your plot with:

```{r}
ggsave("my-plot.png", width = 6, height = 4, dpi = 300)
```