# Interactive Content in Quarto

Quarto supports various forms of interactive content to enrich your document and engage your audience.

------------------------------------------------------------------------

## 6.1 Using htmlwidgets

You can embed interactive widgets from R directly:

```{r}
#| eval: false
library(plotly)
plot_ly(data = iris, x = ~Sepal.Length, y = ~Petal.Length, color = ~Species, type = 'scatter', mode = 'markers')
```

Other useful widgets:

-   `DT::datatable()`

-   `leaflet::leaflet()`

These are rendered as interactive HTML widgets in the output.

## 6.2 Embedding Shiny Apps

You can embed Shiny apps inside a Quarto document:

```{r}
#| context: shiny
#| eval: false
selectInput("var", "Variable", choices = names(mtcars))
renderPlot({
  hist(mtcars[[input$var]])
})
```

> 📌 Requires `runtime: shiny` in the YAML header.

## 6.3 Observable JS Blocks

Use Observable blocks to embed JavaScript-powered interactivity:

```{ojs}
Plot.plot({
  marks: [
    Plot.dotY([1, 2, 4, 8, 16])
  ]
})
```

> Quarto automatically includes Observable runtime when needed.

## 6.4 Code Folding and Reveal

Enable collapsible code blocks:

``` markdown
format:
  html:
    code-fold: true
```

Use `revealjs` format for slideshow-like interaction:

``` markdown
format:
  html:
    code-fold: true
```

## 6.5 User Input + Reactive Output (Shiny)

Use input/output bindings to create small tools or exploratory panels.

```{r}
#| context: shiny
#| eval: false
sliderInput("bins", "Number of Bins", min = 1, max = 50, value = 30)
renderPlot({
  hist(faithful$waiting, breaks = input$bins)
})
```